Data Specialization for a Staged Lambda Calculus

### 2015-02-27 paper/splitting.tex
Quick first pass at 4.4.

\textbf{\bbonem\ term splitting.}  For any $\colmix e A$ which reduces
to $\rtab \xi v$ (via $\redonesym$) , splitting $e$ (via the new
relation~$\splitonesym$) again yields the pair of monostage terms
$\pipeM c l r$. Evaluating $c$ in the first pass produces the pair
$(i,b)$, where the first element $i$ matches the first stage results
contained in $v$.  For example, in the simple case where $\rtab \xi v$
contains entirely first stage computation, such as:
\begin{lstlisting}
2`[]`1` grnd{1+2}`
\end{lstlisting}
the first stage result is the value 3.  The second element of the pair
is the boundary value used as input to the resumer in second pass
evaluation as $\reduce {[b/l]r} {v'}$, where $v'$ is the same as
result produced by second stage computations in $e$.  For example, in
the case where $\rtab \xi v$ contains entirely second stage
computation, such as:
\begin{lstlisting}
2`[yhat |-> 2+3]`1` next{`2`yhat`1`}`
\end{lstlisting}
the second stage result is the value 5. Since in general residual
tables and partial values describe a mixture of both first and second
stage computations, in the subsequent subsection we define an
operation, called {\em masking}, that separates residual tables and
partial values into stage one and stage two results. We then describe
the algorithm for splitting both \bbonem\ terms ($\splitonesym$) and
\bbtwo\ terms ($\splittwosym$).

@@

The splitting algorithm described in the previous subsections is
defined based on the local structure of \lang\ terms.  One artifact of
this design is that splitting generates

For example, the rule for splitting \bbonem\ case terms

% splitting rules are local term driven

% generate a lot of cruft

% for example, case pushes the state of the case through to 2 even if branches have no stage 1 content

% recursive case would be recursive calls

% Rather than try to optimize globally So we use the type system to prove statically there no stage 1 stuff

% skip splitting.

% example is partition in 6.3

In order to be correct, any code at \bbonem\ must be split with the 
pessimistic assumption that it may result in work at the second stage.
When this assumption turns out not to be true---that is, purely monostage code---then 
splitting may produce second stage code which is needlessly costly 
(an example of this will occur in \ref{sec:exampleQS}).
Detecting and optimizing this case is in general a global program analysis,
since functions can be passed around as values.  
So instead of relying on hefty analysis, 
we take the approach of adding enough structure to the input language's type system to
allow the input code to {\em prove} itself to be---in some parts---monostage.

%We find that giving input with \texttt{grnd} annotations is not especially cumbersome in practice.
%Yet even if a language designer desired a different implicit/explicit trade-off,
%then the three-world \lang\ would still be useful as a typed intermediate representation.

>>>>

\textbf{\bbonem\ term splitting.}  For any $\colmix e A$ which reduces
to $\rtab \xi v$ (via $\redonesym$) , splitting $e$ (via the new
relation~$\splitonesym$) again yields the pair of monostage terms
$\pipeM c l r$. Evaluating $c$ in the first pass produces the pair
$(i,b)$, where the first element $i$ matches the first stage results
contained in $v$.  For example, in the simple case where $\rtab \xi v$
contains entirely first stage computation, such as:
\begin{lstlisting}
2`[]`1` grnd{1+2}`
\end{lstlisting}
the first stage result is the value 3.  The second element of the pair
is the boundary value used as input to the resumer in second pass
evaluation as $\reduce {[b/l]r} {v'}$, where $v'$ is the same as
result produced by second stage computations in $e$.  For example, in
the case where $\rtab \xi v$ contains entirely second stage
computation, such as:
\begin{lstlisting}
2`[yhat |-> 2+3]`1` next{`2`yhat`1`}`
\end{lstlisting}
the second stage result is the value 5. Since in general residual
tables and partial values describe a mixture of both first and second
stage computations, in the subsequent subsection we define an
operation, called {\em masking}, that separates residual tables and
partial values into their first and second stage results. We then
describe the algorithm for splitting both \bbonem\ terms
($\splitonesym$) and \bbtwo\ terms ($\splittwosym$).

@@

The splitting algorithm described in the previous subsections operates
on the local structure of \lang\ terms.  One artifact of this design
is that splitting \bbonem\ terms may generate resumers containing
unnecessary logic.  For example, the rule for splitting
\bbonem\ \texttt{case} terms inserts the tag from the \texttt{case}
argument into the boundary value, then decodes this tag in the
resumer.  This logic occurs regardless of whether the terms forming
the branches of the \texttt{case} contain second stage computations.
Worse, if this \texttt{case} appeared in the body of a recursive
function with no other second stage computations, splitting would
generate a resumer with (useless) recursive calls.

Rather than attempt global optimization of the outputs of splitting,
we instead introduce world \bbonep\ and use the type system to
statically prove terms contain no second stage computations.  In these
situations, splitting is trivial---splitting $\pure e$ terms yields a
unit resumer.

The \texttt{partition} function in the \texttt{quickselect} example in
\ref{sec:exampleQS} is an example of such a function that contains no
second stage computations, but, if not defined within a $\pure$
annotation, would cause splitting to generate a recursive resumer.
